Application of Machine Learning to Increase the Efficiency of the Global Search Algorithm for Solving Multicriterial Problems\thanks{This research was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-10-14 2023/NUMTA/Full paper/NUMTA_2023_MCO.tex
fix commas

%
\maketitle              % typeset the header of the contribution
%
	\begin{abstract}
Decision making models described as problems of multicriterial optimization are very complicated for investigation because the property of criteria contradictoriness leads to the notion of the solution as the set of non-dominated parameters (Pareto set). The complexity of these problems increases significantly in the case where criteria are multiextremal.
In the paper a novel algorithm for multicriterial black-box optimization with multiextremal criteria is considered. This algorithm applies ideas of complexity reduction when the initial multicriterial problem is reduced to a set of univariate scalar subproblems on the base of Peano mapping and maximum convolution. For solving univariate problems an information-statistical global optimization algorithm with guaranteed convergence to global optimum is used. As a core of novelty, this algorithm includes in its computational scheme a machine learning procedure with combination of accumulating the information of solved subproblems that allows one to significantly accelerate building the Pareto set.
Effectiveness of the proposed approach is estimated in the representative computational experiment on test sets of multicriterial problems with multiextremal criteria for different dimensions in comparison with several nature-inspired multiobjective optimization algorithms.

@@

The third approach is based on the reduction of multicriterial problem to a family of single-criterion (scalar) optimization subproblems by means of convolution method and solving these subproblems by efficient mathematical programming algorithms \cite{Collette2004,Marler2009,Marler2004,Pardalos2017,Wierzbicki,Ehrgott2005,GergelKozinov2020}. Under corresponding conditions, the solution of the scalar subproblem is a Pareto-optimal point. When changing coefficients of the convolution it is possible to find an approximation of the Pareto set. If the criteria in the initial multiobjective statement are multiextremal, the reduced convolution function is multiextremal as well, and for its optimization it is necessary to apply search algorithms that provide the finding of the global optimum.

@@

The proposed method sequentially implements the concept of complexity reduction. Following this way, the initial MMCO problem is reduced to a family of single-criterion optimization subproblems by means of parameterized convolution. Then each multidimensional subproblem is transformed to an equivalent univariate (one-dimensional) optimization problem formed on the base of Peano-type mapping \cite{Evtushenko2014,Gergel2018,Marler2004,Pardalos2017}. Further, a special global optimization algorithm designed on the ISA platform is applied for solving the univariate problems. Its computational scheme includes two accelerators that decreases significantly number of algorithm's iterations. The first one when solving current one-dimensional problem takes into account results of iterations of all the previous problems solved already. The second mechanism influences the choice of the new iteration point estimating possible closeness of its pre-image in the criteria space to the Pareto set. This estimation is carried out by the method of separation planes, which is widely used in machine learning.

@@

Unfortunately, finding a decision   that provides the minimal values for all partial criteria simultaneously is possible only in exceptional cases because, as a rule, partial criteria are contradictory, when decreasing one criterion leads to increasing the other.

@@

If in the domain $H$ all the criteria 
\begin{equation}
    \label{eq:07}
    \varphi_s(y) > 0, 1 \leq s \leq k,
\end{equation}
then global minimizer of the problem
\begin{equation}
    \label{eq:08}
    W(\mu, y) = V(\mu, y) + \alpha \sum_{s=1}^k \mu_s \varphi_s(y) \to \min, \; y \in H,
\end{equation}
is a Pareto point of the problem (\ref{eq:03}) ($\alpha$  is a sufficiently small parameter) \cite{Wierzbicki,Marler2004}.

@@

However, because of fundamental characteristics of the mapping $y(x)$ the reduced function $g(y(x))$ does not retain Lipschitz property in the case of Lipschitzian partial criteria from (\ref{eq:02}). In this situation the function $g(y(x))$ satisfies the H{\" o}lder condition 
\begin{equation}
    \label{eq:10}
    |g(y(x')) - g(y(x''))| \leq G \left|x'-x''\right|^{1/n}, \; x', x'' \in [0,1],
\end{equation}
where $G>0$ is H{\" o}lder constant. If the function $g(y)$ satisfies the Lipschitz property with a constant $L$ then the function $g(y(x))$ will satisfy the inequality (\ref{eq:10}) with $G=2L\sqrt{n+3}$ \cite{Strongin2003,Sergeyev2013}.

@@

Let the set $M$ contain the finite number $m$ of different weight vectors (\ref{eq:06a}):
\begin{equation}
    \label{eq:11}
    M=(\mu^1, \dots, \mu^m),
\end{equation}
where
\begin{equation}
    \label{eq:12}
    \mu^i=(\mu^i_k, \dots, \mu^i_k), \; 1 \leq i \leq m.
\end{equation}

@@

It should be noted that if the value (\ref{eq:17}) has been already calculated and the vector $\Omega^i(x^j)$ has been stored, then evaluation of another function with its weight coefficient set $M'=(\mu'^1, \dots, \mu'^k)$ at the point $x^j$ does not require the recalculation the partial criteria (\ref{eq:02}) because the value $z^{l, j}$ can be easily computed according to (\ref{eq:17}).

@@

In all the stages after the first one $(i>1)$, the trial coordinates obtained in all previous stages are taken as points of initial trials at current stage. As the matrix $A^{i-1}_q$ contains values of all partial criteria at these points it is easy to recalculate values $z^{i,1}, z^{i,2}, \dots, z^{i,p}$ of function $f^i(x)$ according to (\ref{eq:17}) and to build the updated matrix $A^i_q$.

@@

\textbf{Rule 4.} In the criteria space on the base of the marked up matrix $A^i_p$ a separating hyperplane $L_p$ is constructed applying logistic regression \cite{Yu2011} as efficient tool of machine learning.

@@

The data of the table show that AGS-SP and AGS-R solve the problem (\ref{eq:28}) more efficient than evolutionary methods because while spending the same number of trials (and even less as AGS-SP) they provide the  better HV index and find more points in Pareto set approximation. If to compare AGS-SP and AGS-R with each other, then AGS-SP has an advantage due to HV indices being the same, but AGS-SP was ``faster'' as it executed fewer trials.

>>>>

%
\maketitle              % typeset the header of the contribution
%
	\begin{abstract}
Decision making models described as problems of multicriterial optimization are very complicated for investigation, because the property of criteria contradictoriness leads to the notion of the solution as the set of non-dominated parameters (Pareto set). The complexity of these problems increases significantly in the case where criteria are multiextremal.
In the paper a novel algorithm for multicriterial black-box optimization with multiextremal criteria is considered. This algorithm applies ideas of complexity reduction when the initial multicriterial problem is reduced to a set of univariate scalar subproblems on the base of Peano mapping and maximum convolution. For solving univariate problems an information-statistical global optimization algorithm with guaranteed convergence to global optimum is used. As a core of novelty, this algorithm includes in its computational scheme a machine learning procedure with combination of accumulating the information of solved subproblems that allows one to significantly accelerate building the Pareto set.
Effectiveness of the proposed approach is estimated in the representative computational experiment on test sets of multicriterial problems with multiextremal criteria for different dimensions, in comparison with several nature-inspired multiobjective optimization algorithms.

@@

The third approach is based on the reduction of multicriterial problem to a family of single-criterion (scalar) optimization subproblems by means of convolution method, and solving these subproblems by efficient mathematical programming algorithms \cite{Collette2004,Marler2009,Marler2004,Pardalos2017,Wierzbicki,Ehrgott2005,GergelKozinov2020}. Under corresponding conditions, the solution of the scalar subproblem is a Pareto-optimal point. When changing coefficients of the convolution it is possible to find an approximation of the Pareto set. If the criteria in the initial multiobjective statement are multiextremal, the reduced convolution function is multiextremal as well, and for its optimization it is necessary to apply search algorithms that provide the finding of the global optimum.

@@

The proposed method sequentially implements the concept of complexity reduction. Following this way, the initial MMCO problem is reduced to a family of single-criterion optimization subproblems by means of parameterized convolution. Then each multidimensional subproblem is transformed to an equivalent univariate (one-dimensional) optimization problem formed on the base of Peano-type mapping \cite{Evtushenko2014,Gergel2018,Marler2004,Pardalos2017}. Further, a special global optimization algorithm designed on the ISA platform is applied for solving the univariate problems. Its computational scheme includes two accelerators that significantly decrease the number of algorithm's iterations. The first one when solving current one-dimensional problem takes into account results of iterations of all the previous problems solved already. The second mechanism influences the choice of the new iteration point estimating possible closeness of its pre-image in the criteria space to the Pareto set. This estimation is carried out by the method of separation planes, which is widely used in machine learning.

@@

Unfortunately, finding a decision   that provides the minimal values for all partial criteria simultaneously is possible only in exceptional cases because, as a rule, partial criteria may be contradictory, i.e. when decreasing one criterion leads to increasing the other.

@@

If in the domain $H$ all the criteria satisfy
\begin{equation}
    \label{eq:07}
    \varphi_s(y) > 0, 1 \leq s \leq k,
\end{equation}
then global minimizer of the problem
\begin{equation}
    \label{eq:08}
    W(\mu, y) = V(\mu, y) + \alpha \sum_{s=1}^k \mu_s \varphi_s(y) \to \min, \; y \in H,
\end{equation}
is a Pareto point of the problem (\ref{eq:03}) ($\alpha$  is a sufficiently small parameter) \cite{Wierzbicki,Marler2004}.

@@

However, because of fundamental characteristics of the mapping $y(x)$, the reduced function $g(y(x))$ does not retain Lipschitz property in the case of Lipschitzian partial criteria from (\ref{eq:02}). In this situation the function $g(y(x))$ satisfies the H{\" o}lder condition 
\begin{equation}
    \label{eq:10}
    |g(y(x')) - g(y(x''))| \leq G \left|x'-x''\right|^{1/n}, \; x', x'' \in [0,1],
\end{equation}
where $G>0$ is H{\" o}lder constant. If the function $g(y)$ satisfies the Lipschitz property with a constant $L$ then the function $g(y(x))$ will satisfy the inequality (\ref{eq:10}) with $G=2L\sqrt{n+3}$ \cite{Strongin2003,Sergeyev2013}.

@@

Let the set $M$ contain the finite number $m$ of different weight vectors (\ref{eq:06a}):
\begin{equation}
    \label{eq:11}
    M=(\mu^1, \dots, \mu^m),
\end{equation}
where
\begin{equation}
    \label{eq:12}
    \mu^i=(\mu^i_1, \dots, \mu^i_k), \; 1 \leq i \leq m.
\end{equation}

@@

It should be noted that if the value (\ref{eq:17}) has been already calculated and the vector $\Omega^i(x^j)$ has been stored, then evaluation of another function with its weight coefficient set $M'=(\mu'^1, \dots, \mu'^k)$ at the point $x^j$ does not require the recalculation of the partial criteria (\ref{eq:02}) because the value $z^{l, j}$ can be easily computed according to (\ref{eq:17}).

@@

In all the stages after the first one $(i>1)$, the trial coordinates obtained in all previous stages are taken as points of initial trials at current stage. As the matrix $A^{i-1}_q$ contains values of all partial criteria at these points, it is easy to recalculate values $z^{i,1}, z^{i,2}, \dots, z^{i,p}$ of function $f^i(x)$ according to (\ref{eq:17}) and to build the updated matrix $A^i_q$.

@@

\textbf{Rule 4.} In the criteria space on the base of the marked up matrix $A^i_p$, a separating hyperplane $L_p$ is constructed applying logistic regression \cite{Yu2011} as efficient tool of machine learning.

@@

The data of the table shows that AGS-SP and AGS-R solve the problem (\ref{eq:28}) more efficiently than evolutionary methods because while spending the same number of trials (and even less as AGS-SP) they provide the  better HV index and find more points in Pareto set approximation. If to compare AGS-SP and AGS-R with each other, then AGS-SP has an advantage due to HV indices being the same, but AGS-SP was ``faster'' as it executed fewer trials.
